---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# expirest

<!-- badges: start -->
<!-- badges: end -->

The package supports shelf life estimation for chemically derived medicines, either following the standard method proposed by the International Council for Harmonisation (ICH), in quality guideline Q1E Evaluation of Stability Data or following the worst-case scenario consideration (what-if analysis) described in the Australian Regulatory Guidelines for Prescription Medicines (ARGPM), guidance on Stability testing for prescription medicines.


## Installation

A stable version of `expirest` can be installed from CRAN:

``` r
# install.packages("expirest")
```

The development version is available from [GitHub](https://github.com/piusdahinden/expirest) by:

``` r
# install.packages("devtools")
devtools::install_github("piusdahinden/expirest")
```


## Examples

### Example 1
This is a basic example which shows you how to solve a common problem using a data set containing the moisture stability data (% (w/w)) of three batches obtained over a 24 months period of a drug product. A total of *n* = 33 independent measurements are available (corresponding to data shown in Table XIII in LeBlond et al. (2011).

``` {r example-1, message = FALSE}
library(expirest)

# Data frame
str(exp3)

# Perform what-if shelf life estimation (wisle) and print a summary
res1 <- expirest_wisle(
      data = exp3, response_vbl = "Moisture", time_vbl = "Month",
      batch_vbl = "Batch", rl = 3.00, rl_sf = 3, sl = c(0.5, 4.5),
      sl_sf = c(1, 2), srch_range = c(0, 500), alpha = 0.05,
      alpha_pool = 0.25, xform = c("no", "no"), shift = c(0, 0),
      sf_option = "tight", ivl = "confidence", ivl_type = "one.sided",
      ivl_side = "upper")

class(res1)
summary(res1)

# Prepare graphical representation
ggres1 <- plot_expirest_wisle(
      model = res1, rl_index = 1, response_vbl_unit = "% (w/w)",
      x_range = NULL, y_range = c(0.2, 5.2), scenario = "standard",
      mtbs = "verified", plot_option = "full", ci_app = "ribbon")
class(ggres1)
plot(ggres1)
```


### Example 2
The model type in *Example 1* was *common intercept / common slope* (cics). The model type in this example is *different intercept / different slope* (dids). A data set containing the potency stability data (in \% of label claim (LC)) of five batches of a drug product obtained over a 24 months period is used. A total of *n* = 53 independent measurements are available (corresponding to data shown in Tables IV, VI and VIII in LeBlond et al. (2011).

``` {r example-2-1, message = FALSE}
# Data frame
str(exp1)

# Perform what-if shelf life estimation (wisle) and print a summary
res2_1 <- expirest_wisle(
      data = exp1[exp1$Batch %in% c("b4", "b5", "b8"), ],
      response_vbl = "Potency", time_vbl = "Month", batch_vbl = "Batch",
      rl = c(98.0, 98.5, 99.0), rl_sf = rep(2, 3), sl = 95, sl_sf = 2,
      srch_range = c(0, 500), alpha = 0.05, alpha_pool = 0.25,
      xform = c("no", "no"), shift = c(0, 0), sf_option = "tight",
      ivl = "confidence", ivl_type = "one.sided", ivl_side = "lower")

summary(res2_1)

# Prepare graphical representation
ggres2_1 <- plot_expirest_wisle(
      model = res2_1, rl_index = 2, response_vbl_unit = "%", x_range = NULL,
      y_range = c(93, 107), scenario = "standard", mtbs = "verified",
      plot_option = "full", ci_app = "ribbon")
class(ggres2_1)
plot(ggres2_1)
```

Note that if the model type is dids, then by default the results are plotted that are obtained by fitting individual regression lines to each batch. To get a plot of the dids model with pooled mean square error (pmse) instead, use the setting `mtbs = "dids.pmse"`.

``` {r example-2-2, message = FALSE}
# Prepare graphical representation of the dids.pmse model
ggres2_2 <- plot_expirest_wisle(
      model = res2_1, rl_index = 2, response_vbl_unit = "%", x_range = NULL,
      y_range = c(93, 107), scenario = "standard", mtbs = "dids.pmse",
      plot_option = "full", ci_app = "ribbon")
plot(ggres2_2)
```

The examples above were performed with three batches. The assessment can be performed with any numbers of batches, even with just one batch. If the assessment is performed with a single batch, the results are reported as dids model while the results for all other models are NA. As an example, the following example was performed with batch b4.

``` {r example-2-3, message = FALSE}
# Perform what-if shelf life estimation (wisle) and print a summary
res2_2 <- expirest_wisle(
      data = exp1[exp1$Batch == "b4", ],
      response_vbl = "Potency", time_vbl = "Month", batch_vbl = "Batch",
      rl = c(98.0, 98.5, 99.0), rl_sf = rep(2, 3), sl = 95, sl_sf = 2,
      srch_range = c(0, 500), alpha = 0.05, alpha_pool = 0.25,
      xform = c("no", "no"), shift = c(0, 0), sf_option = "tight",
      ivl = "confidence", ivl_type = "one.sided", ivl_side = "lower")

summary(res2_2)

# Prepare graphical representation
ggres2_2 <- plot_expirest_wisle(
      model = res2_2, rl_index = 2, response_vbl_unit = "%", x_range = NULL,
      y_range = c(93, 107), scenario = "standard", mtbs = "verified",
      plot_option = "full", ci_app = "ribbon")
plot(ggres2_2)
```


### Example 3
The assessments in *Example 1* and *Example 2* were made with the option `sf_option = "tight"`. The parameter `sf_option` affects the shelf life (*sl*) and release limits (*rl*). The setting `sf_option = "tight"` means that of the *sl* and *rl* limits only the number of significant digits specified by the parameters `rl_sf` and `sl_sf`, respectively, are taken into account. If `sf_option = "loose"`, then the limits are widened, i.e. four on the order of the last significant digit + 1 is added if `ivl_side = "upper"` or five on the order of the last significant digit + 1 is subtracted if `ivl_side = "lower"`. For illustration, the following example reproduces *Example 2* with the option `sf_option = "loose"`.

``` {r example-3-1, message = FALSE}
# Perform what-if shelf life estimation (wisle) and print a summary
res3 <- expirest_wisle(
      data = exp1[exp1$Batch %in% c("b4", "b5", "b8"), ],
      response_vbl = "Potency", time_vbl = "Month", batch_vbl = "Batch",
      rl = c(98.0, 98.5, 99.0), rl_sf = rep(3, 3), sl = 95, sl_sf = 2,
      srch_range = c(0, 500), alpha = 0.05, alpha_pool = 0.25,
      xform = c("no", "no"), shift = c(0, 0), sf_option = "loose",
      ivl = "confidence", ivl_type = "one.sided", ivl_side = "lower")

summary(res3)

# Prepare graphical representation
ggres3 <- plot_expirest_wisle(
      model = res3, rl_index = 2, response_vbl_unit = "%", x_range = NULL,
      y_range = c(93, 107), scenario = "standard", mtbs = "verified",
      plot_option = "full", ci_app = "ribbon")
plot(ggres3)
```


## Literature
LeBlond, D., Griffith, D. and Aubuchon, K. Linear Regression 102: Stability
Shelf Life Estimation Using Analysis of Covariance. *J Valid Technol*
(2011) **17**(3): 47-68.


## Contact
[Pius Dahinden](https://github.com/piusdahinden), Tillotts Pharma AG

---
